---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
<p> The dataset is a comma-separated-value (CSV) file with a total of **17,568** observations</p>

<p> 1. To load the data </p>
<ul>
  <li> unzip the data file (activity.zip) </li>
  <li> read the data using read.csv </li>
</ul>

<p> 2. The results of the loading process is as follows: </p> 

```{r readData, echo=TRUE}

library(dplyr)
library(data.table)
library(knitr)
library(ggplot2)

setwd("~/Coursera/RepData_PeerAssessment1")
unzip("activity.zip")
data <- read.csv("activity.csv")

str(data)

```

<hr></hr>

## What is mean total number of steps taken per day?

```{r Total_Steps_Day}

## Dataset: total number of steps taken per day
total_steps_day <- summarize(group_by(data, date), tot_steps = sum(steps, na.rm=TRUE))

## Calculate the mean and median total number of steps taken per day
mean_steps <- mean(total_steps_day$tot_steps, na.rm=TRUE)
median_steps <- median(total_steps_day$tot_steps, na.rm=TRUE)

## histogram of the total number of steps taken each day
qplot(total_steps_day$tot_steps, 
      geom="histogram",
      binwidth = 1000,
      ylim = c(0, 20),
      main = "Total Number of Steps Taken each day",
      xlab = "Total Steps per day")

summary(total_steps_day) #check mean from summary

```

<p> 1. Total number of steps taken per day (calculation):
<ul>
    <li> Mean: **`r format(mean_steps, big.mark=",")`** </li>
    <li> Median: **`r format(median_steps, big.mark=",")`** </li>
</ul>
</p>
<hr></hr>

## What is the average daily activity pattern?

```{r Average_Daily_Activity_Patterns}

## Dataset: average number of steps across all the days in the dataset per 5-minute interval
avg_steps_interval <- summarize(group_by(data, interval), avgsteps = mean(steps, na.rm=TRUE))


## Determine the interval with the maximum number of steps
max_interval <- subset(avg_steps_interval, avgsteps == max(avgsteps))

##  Time series plot (i.e. type = "l") of the 5-minute interval (x-axis) 
##  and the average number of steps taken, averaged across all days (y-axis)
with (avg_steps_interval, 
    plot(interval, avgsteps, 
           type="l",
           main = "Average Daily Activity Pattern \n averaged at 5-minute Intervals across all days",                       
           xlab = "Interval",
           ylab = "Average number of steps"))


```

<p> 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps? **`r max_interval$interval`** </p>

<hr></hr>


## Imputing missing values
#### Strategy: *replace NAs with mean for that 5-minute interval*

```{r Missing_Values}

## Calculate number of missing values
num_missing_vals <- sum(is.na(data$steps))

## Dataset: Original dataset merged with data from 5-minute interval dataset.
adj_data <- data.table(merge(data, avg_steps_interval, by="interval"))

## Make a copy step data for adjustment
adj_data$adj_steps <- adj_data$steps

## Replace NAs (step) with 5-minute interval average
adj_data[is.na(adj_data$adj_steps), adj_steps := as.integer(avgsteps)]

## Calculate total steps taken per day adjusted and with NA
total_steps_day <- summarize(group_by(adj_data, date), 
                             tot_steps_adj = sum(adj_steps), 
                             tot_steps_na = sum(steps, na.rm=TRUE))

## Calculate the mean and median total number of steps taken per day
mean_steps <- mean(total_steps_day$tot_steps_adj)
median_steps <- median(total_steps_day$tot_steps_adj)

## histogram of the total number of steps taken each day
qplot(total_steps_day$tot_steps_adj, 
      geom="histogram",
      binwidth = 1000,
      ylim = c(0, 20),
      main = "Total number of steps taken each day \n (Missing Data imputed: tot_steps_adj)",
      xlab = "Total Steps per day")

tbl <- total_steps_day[total_steps_day$tot_steps_na == 0, ]

```

<p> 1. Total number of missing values in the dataset: **`r num_missing_vals`** </p>

<p> 2. Strategy: replace NAs with mean for that 5-minute interval </p>

<p> 4. Total number of steps taken per day: 
<ul>
      <li> Mean: **`r format(mean_steps, big.mark=",")`** </li>      
      <li> Median: **`r format(median_steps, big.mark=",")`** </li>
</ul></p>


<p> 4. The impact of imputing missing data on the estimates of the total daily number of steps is: 
<ul>
    <li> the **(8) days** when no data is recorded are now plotted as 10641 (count is higher) instead of 0 </li>
    <li> the mean and median scores are higher </li>
</ul></p>

`r kable(tbl)`


## Are there differences in activity patterns between weekdays and weekends?

```{r}

## new factor variable in the dataset with two levels -- "weekday" and "weekend" 
adj_data$day_wk <- as.factor(as.POSIXlt(adj_data$date)$wday)

adj_data$wk_end <- as.factor("Weekday")
adj_data[as.character(adj_data$day_wk) %in% c("5":"6"), wk_end := as.factor("Weekend")]


ggplot(adj_data, aes(interval, adj_steps)) + geom_line() + facet_wrap(~wk_end, nrow=2) + labs(title = "Average Daily Activity Pattern \n averaged at 5-minute Intervals") + ylab("Number of steps") + theme(legend.position = "top")


```
